---
title: "Validate and load electronic health records"
author: "Peter Dutey"
date: '`r format(Sys.Date(), "%d %B %Y")`'
output: 
  rmarkdown::html_vignette:
   toc: true 
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{Validate and load electronic health records}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(Ramses)
```


# Introduction

This vignette demonstrates how to clean and validate hospital records and load them into a data warehouse.

Ramses is provided simulated datasets consisting of records of `r formatC(length(unique(inpatient_episodes$spell_id)), big.mark = ",")` hospitalisations for `r formatC(length(unique(inpatient_episodes$patient_id)), big.mark = ",")` patients, complete with prescriptions, diagnoses and care episodes records.


# Setup

First, packages must be loaded.

```{r setup, warning=F, message=F}
library(Ramses)
library(dplyr)    # to facilitate data manipulations
library(magrittr) # for the pipe operator `%>%`
```

Second, a relational database server is needed. In this vignette, we will create a small \href{https://www.sqlite.org}{SQLite 3} database on the local hard drive. This database will behave in a similar way as a large Microsoft SQL Server or MySQL relational database. The database connection information is stored in an object entitled `ramses_db` which must be provided to most functions in the Ramses package.

```{r}
ramses_db <- connect_db_local("ramses-db.sqlite")
```

We can now begin preparing records and loading them into the database. This involves three steps:

1. *Transform*: adding metadata and transforming records to conform with Ramses data validation constraints
2. *Validate*: using dedicated functions (eg: `validate_inpatient_episodes()`) to verify that all data conform to Ramses data validation constraints
3. *Load*: loading records into a SQL database.

# Medication records

## Drug prescriptions

We begin with prescription records (but you can choose to start with other records type if you prefer).

```{r print_prescriptions}
str(drug_prescriptions)
```

Prescriptions records must conform to data validation requirements listed in the documentation for `validate_prescriptions()`

We first map drug concepts thanks to the `as.ab()` function provided by the AMR package. This function takes a character variable as input and returns a three-letter code EARS-Net code. For more information, refer to the documentation for the [AMR::as.ab()] `as.ab()` function. Some manual editing may be required when drugs are not successfully mapped, as shown in the example below.

```{r map_drug_failing}
# attempting to map drug name using AMR package
drug_prescriptions$drug_id <- AMR::as.ab(drug_prescriptions$tr_DESC)
```

```{r map_drug}
# editing drug names
drug_prescriptions$drug_id <- gsub("Vancomycin protocol", 
                                   "Vancomycin",
                                   drug_prescriptions$tr_DESC)
# mapping drug name using AMR package
drug_prescriptions$drug_id <- AMR::as.ab(drug_prescriptions$drug_id)
drug_prescriptions$drug_name <- AMR::ab_name(drug_prescriptions$drug_id)
# recoding route of administration
drug_prescriptions <- mutate(drug_prescriptions, 
  ATC_route = 
    case_when(
      route %in% c(NULL) ~ "Implant", 
      route %in% c("NEB", "INHAL") ~ "Inhal", 
      route %in% c("TOP", "EYE", "EYEL", "EYER", "EYEB", 
                   "EAR", "EARL", "EARR", "EARB") ~ "Instill", 
      route %in% c("NASAL", "NOST", "NOSTL", "NOSTR", "NOSTB") ~ "N", 
      route %in% c("ORAL", "NAS", "PEG") ~ "O", 
      route %in% c("IV", "IVB", "IVI", "IMI", "IT", "IVT") ~ "P", 
      route %in% c("PR") ~ "R", 
      route %in% c("BUCC", "SB", "OROM", "SUBL") ~ "SL", 
      route %in% c("SC", "ID") ~ "TD", 
      route %in% c("PV") ~ "V", 
      TRUE ~ "NA_character_"
    ))
```

With the AMR package, it is possible to infer important properties of the drug:

* its ATC class and group, which can be useful for instance in order identify antifungals (group 'Antifungals/antimycotics')
* Defined Daily Doses (DDDs). 

Working out the DDD of the prescription is somewhat more complex. The reference DDD is for the main substance. In the case of amoxicillin/clavulanic acid for example, the DDD (0.5g) refers to the amoxicillin. A 625mg tablet of Augmentin (c) contains 500mg of amoxicillin and 125mg of clavulanic acid. The reference DDD is `paste0(AMR::ab_ddd("amoxicillin/clavulanic acid", "oral"), AMR::ab_ddd("amoxicillin/clavulanic acid", "oral", unit = T))` = `r paste0(ab_ddd("amoxicillin/clavulanic acid", "oral"), ab_ddd("amoxicillin/clavulanic acid", "oral", unit = T))`. A prescription of Augmentin 625mg TDS (three times a day) therefore amounts to 1.5g of amoxicillin per day, which is 1.5 DDDs. Co-trimoxazole is a compound of 5 parts of sulfamethoxazole to 1 part of trimethoprim. The DDD of sulfamethoxazole is 2g while the DDD of trimethoprim is 400mg: with this ratio of 5 to 1, DDDs calculated on the basis of the dosage of sulfamethoxazole will be valid for co-trimoxazole prepared with 5 parts to 1 part.

These just some examples. In order to obtain the correct DDDs for compound medications, it is recommended to map data to a suitable dictionary of medicines such as the [RxNorm](http://bioportal.bioontology.org/ontologies/RXNORM) or the [NHS Dictionary of Drugs and Medical Devices (dm+d)](https://www.nhsbsa.nhs.uk/pharmacies-gp-practices-and-appliance-contractors/dictionary-medicines-and-devices-dmd) and extract the relevant strengths.

In this example, however, we use a look-up table for the strength of compound medications. Another lookup table `reference_drug_frequency` available within Ramses is also used to convert the `frequency` character variable into a numeric variable indicating the daily frequency of administration. 

```{r compute_ddd}
# extracting the ATC code (eg J01CR02) group (eg Beta-lactam antibacterials, penicillins)
drug_prescriptions$ATC_code <- AMR::ab_atc(drug_prescriptions$drug_id)
drug_prescriptions$ATC_group <- AMR::ab_atc_group1(drug_prescriptions$drug_id)
# removing antifungal drug, if applicable
drug_prescriptions <- filter(drug_prescriptions, 
                             ATC_group != "Antimycotics for systemic use")

# prepare DDD extraction
compound_strength_lookup <- data.frame(list(
  drug_id = c("AMC", "AMC", "TZP", "SMX"),
  route = c("oral", "oral", "oral", "oral"),
  dose = c(625, 1.2, 4.5, 480),
  units = c("mg", "g", "g", "mg"),
  strength = c(500, 1, 4, 400),
  basis_of_strength = c("AMX", "AMX", "PIP", "SMX")
), stringsAsFactors = F)

drug_prescriptions <- merge(drug_prescriptions, 
                            compound_strength_lookup, 
                            all.x = T)
drug_prescriptions <- drug_prescriptions %>% 
  mutate(strength = if_else(is.na(strength), dose, strength),  
         basis_of_strength = if_else(is.na(basis_of_strength),
                                     as.character(drug_id),
                                     basis_of_strength))

drug_prescriptions <- merge(drug_prescriptions, 
                            reference_drug_frequency, by = "frequency", all.x = T)

# computing the prescription DDD the reference DDD from the ATC
drug_prescriptions <- drug_prescriptions %>% 
  mutate(daily_dose = strength * daily_frequency) %>% 
  mutate(DDD = compute_DDDs(
    ATC_code = AMR::ab_atc(basis_of_strength),
    ATC_administration = ATC_route,
    dose = daily_dose,
    unit = units))


```


```{r}
drug_prescriptions <- drug_prescriptions %>% 
  transmute(patient_id,
            prescription_id,
            drug_id,
            drug_name = drug_name,
            drug_display_name = drug_name,
            ATC_code,
            ATC_group,
            ATC_route,
            authoring_date,
            prescription_start,
            prescription_end,
            prescription_status = "completed",
            prescription_context = "inpatient",
            dose,
            unit = units,
            route,
            frequency,
            daily_frequency,
            DDD)
```

## Drug administrations

An identical process is followed for drug administration, with the difference that DDDs do not rely on a daily frequency of administration variable

```{r prepare_drug_admin}
drug_administrations$drug_id <- gsub("Vancomycin protocol", 
                                     "Vancomycin", 
                                     drug_administrations$tr_DESC)
drug_administrations$drug_id <- AMR::as.ab(drug_administrations$drug_id)
drug_administrations$drug_name <- AMR::ab_name(drug_administrations$drug_id)
# recoding route of administration
drug_administrations <- mutate(drug_administrations, 
  ATC_route = 
    case_when(
      route %in% c(NULL) ~ "Implant", 
      route %in% c("NEB", "INHAL") ~ "Inhal", 
      route %in% c("TOP", "EYE", "EYEL", "EYER", "EYEB", 
                   "EAR", "EARL", "EARR", "EARB") ~ "Instill", 
      route %in% c("NASAL", "NOST", "NOSTL", "NOSTR", "NOSTB") ~ "N", 
      route %in% c("ORAL", "NAS", "PEG") ~ "O", 
      route %in% c("IV", "IVB", "IVI", "IMI", "IT", "IVT") ~ "P", 
      route %in% c("PR") ~ "R", 
      route %in% c("BUCC", "SB", "OROM", "SUBL") ~ "SL", 
      route %in% c("SC", "ID") ~ "TD", 
      route %in% c("PV") ~ "V", 
      TRUE ~ "NA_character_"
    ))
drug_administrations$ATC_code <- AMR::ab_atc(drug_administrations$drug_id)
drug_administrations$ATC_group <- AMR::ab_atc_group1(drug_administrations$drug_id)
drug_administrations <- filter(drug_administrations, 
                               ATC_group != "Antimycotics for systemic use")

drug_administrations <- merge(drug_administrations, compound_strength_lookup, all.x = T)
drug_administrations <- drug_administrations %>% 
  mutate(strength = if_else(is.na(strength), dose, strength),
         basis_of_strength = if_else(is.na(basis_of_strength),
                                     as.character(drug_id),
                                     basis_of_strength))

drug_administrations <- drug_administrations %>% 
  mutate(DDD = compute_DDDs(
    ATC_code = AMR::ab_atc(basis_of_strength),
    ATC_administration = ATC_route,
    dose = dose,
    unit = units 
  )) 

# Create an identifier
# Note: An alternative is encryption using openssl::sha256()
      
drug_administrations$administration_id <- drug_administrations %>%
  dplyr::group_indices(
    patient_id,
    drug_id,
    route,
    dose,
    units,
    administration_date)

drug_administrations <- drug_administrations %>% 
    transmute(
      patient_id,
      administration_id = as.character(administration_id),
      prescription_id,
      administration_text = paste0(
          drug_name, " ", route, " ", dose, units),
      drug_id,
      drug_name,
      drug_display_name = drug_name,
      ATC_code,
      ATC_group,
      ATC_route,
      dose,
      unit = units,
      route,
      administration_date,
      administration_status = "completed",
      DDD
    )
```


This data frame can now be validated, before being loaded into the database.

```{r load_rx}
validate_administrations(drug_administrations)

load_medications(conn = ramses_db, 
                 prescriptions = drug_prescriptions, 
                 administrations = drug_administrations, 
                 overwrite = TRUE)
```

# Inpatient care records

## Admissions

We now turn to records of admissions, discharges and transfers. Ramses is designed around the notion of *hospital spells* identified by variable `spell_id`, which correspond to the window from the time a patient is admitted to the time a patient is discharged. 

In addition, Ramses use subdivisions of *hospital spells* called *episodes of care*, which designate a continuous period of inpatient care under the responsibility of a department, specialty, or clinical team. This corresponds to the 'finished consultant episode' in the UK NHS Admitted Patient Care data model [@Herbert2017]. All *hospital spells* consist of at least one *episode of care*.
Episodes of care are used to calculate bed-days as well as rates of infection, prescribing and other stewardship metrics for individal specialties and teams of clinicians. 

Hospital spells and episodes of care are recorded in a single table in which every row corresponds to one episode of care.

```{r}
str(inpatient_episodes)
```

Dates of admissions and episode start/end must be valid:

* there must be no overlap between admissions or between episodes
* there must be no gap between episodes in an admission.

```{r}
validate_inpatient_episodes(episodes = inpatient_episodes)
```

Optionally, Ramses can use data on ward movements. These records must be validated together with episodes of care

```{r}
validate_inpatient_episodes(episodes = inpatient_episodes, 
                            wards = inpatient_wards)
```

The data can now be loaded into the database.

```{r}
load_inpatient_episodes(conn = ramses_db,
                        episodes_data = Ramses::inpatient_episodes,
                        wards_data =  Ramses::inpatient_wards)
```

## Diagnosis codes (ICD-10)

In order to exploit clinical diagnoses coded by doctors during hospital admissions, Ramses relies on a range of look up tables which must be built to suit the precise ICD-10 version used by the data. Ramses uses three types of look up tables:

* an ICD-10 reference table containing all ICD-10 codes, their full-text description, and chapter headings
* a categorisation based on the AHRQ Clinical Classification Software (CCS) or its Revised version (CCSR)
* a classification of infections and whether antibiotics are commonly indicated to treat them, adapted from @Hashimoto2020

## ICD-10 reference table

This example uses the `Ramses::inpatient_diagnoses` example dataset. First, we download the ICD-10-CM reference look up table

```{r}
icd10cm <- download_icd10cm()
str(icd10cm)
```

We verify that all ICD-10 codes in the dataset can be mapped to the reference table.
The first validation fails, because some of the `icd_code` in `inpatient_diagnoses` are missing (all diagnoses that are not infections). These are removed before validation is attempted again. In the present example, the dataset is simulated using ICD-10 5th Edition, rather than ICD-10-CM. The match is therefore not complete on the second validation attempt. Although the validation produces a warning, it passes (returns `TRUE`). This means it will not prevent loading the data into the warehouse.

```{r diagnoses_first_valid_attempt}
validate_inpatient_diagnoses(inpatient_diagnoses, icd10cm)
```
```{r diagnoses_second_valid_attempt}
inpatient_diagnoses <- filter(inpatient_diagnoses, !is.na(icd_code))
validate_inpatient_diagnoses(inpatient_diagnoses, icd10cm)
```


Next, we classify the reference look up table to other classifications:

- \link[common infections and indications]{antibiotic_icd_indications} for antibiotic prescribing
- Charlson Comorbidity Index weights
- Clinical Classifications Software (CCS)
- Clinical Classifications Software Refined (CCSR)

Although this is done automatically when loading diagnoses into the warehouse, this step can be useful to first check that the ICD-10 reference table is successfully mapped to other classifications. 

```{r}
icd_infection_lkup <- icd10cm %>% 
  select(icd_code) %>% 
  map_infections_abx_indications(df = ., icd_column = "icd_code")

comorbidity_lkup <- icd10cm %>% 
  select(icd_code) %>% 
  map_charlson_comorbidities(df = ., icd_column = "icd_code")

ccsr_lkup <- icd10cm %>% 
  select(icd_code) %>% 
  map_ICD10_CCSR(df = ., icd_column = "icd_code")

ccs_lkup <- icd10cm %>% 
  select(icd_code) %>% 
  map_ICD10_CCS(df = ., icd_column = "icd_code")

head(icd_infection_lkup)
```


Data may now be loaded into the database. Diagnoses and their reference look-up are loaded in one step

```{r load_diagnoses}
load_inpatient_diagnoses(conn = ramses_db, 
                         diagnoses_data = inpatient_diagnoses, 
                         diagnoses_lookup = icd10cm, 
                         overwrite = TRUE)
```

# Investigation records

## Microbiology cultures and sensitivity 

A small sample of records of microbial isolates is provided along with susceptibility testing results:

```{r show_inpatient_microbiology}
str(Ramses::inpatient_microbiology)
```

Before validating and loading these records into the database, some transformations 
will be required to create three separate data frames:

1. `specimens` contains one row per microbial sample sent to the laboratory
2. `isolates` contains zero, one, or multiple rows for every specimen -- as many rows as the number of microorganisms isolated from the specimen, whatever the technique (eg mass spectrometry, PCR, bacterial culture)
3. `susceptibility` contains results of susceptibility testing: one row per substance tested.

For this, we create a list of data frames named `microbiology_data`, and commence with cleaning the raw data. We employ the [AMR](https://cran.r-project.org/web/packages/AMR/index.html) package to code microorganism species names (`AMR::as.mo()`) and antimicrobial names (`AMR::as.ab`). 

```{r}
microbiology_data <- list()

# Start by cleaning the raw data
microbiology_data$raw <- inpatient_microbiology
microbiology_data$raw <- microbiology_data$raw %>% 
  mutate(organism_display_name = if_else(organism_display_name == "No growth",
                                         NA_character_,
                                         organism_display_name)) %>% 
  mutate(organism_code = AMR::as.mo(organism_display_name),
         drug_id = AMR::as.ab(drug_display_name)) %>% 
  mutate(organism_name = AMR::mo_name(organism_code),
         drug_name = AMR::ab_name(drug_id))
```


As for specimen types (eg bood, bronchoalveolar lavage fluid, etc.), they must be coded using [SNOMED CT specimen concepts](https://snowstorm.ihtsdotools.org/?perspective=full&conceptId1=123038009&edition=MAIN/2020-07-31&release=&languages=en). A complete list of concepts from the SNOMED CT International Edition can be found in `reference_specimen_type`.

Susceptibility testing results in the example dataset are already interpreted as Resistant/Increased exposure/Susceptible (RSI). Where needed, minimum inhibitory concentrations can be interpreted as RSI using the `as.rsi()` function provided by the AMR package.

If possible, use the most specific concept available (eg `309176002 | Bronchial brushings sample (specimen) |`) over a more generic term (eg `258415003 | Biopsy sample (specimen) |`). Specific concepts can always be reclassified to higher-level concepts using SNOMED CT inference (in this case, not just `258415003 | Biopsy sample (specimen) |`, but also `258603007 | Respiratory sample (specimen) |`). The [snomedizer](https://snomedizer.web.app/) package can be used to query SNOMED CT concept parents and descendants.

```{r code_specimen_types}
microbiology_data$raw <- microbiology_data$raw %>% 
  mutate(specimen_type_code = case_when(
    specimen_type_display == "Blood Culture" ~ 
      "446131002", # Blood specimen obtained for blood culture
    specimen_type_display == "Faeces" ~ 
      "119339001", # Stool specimen
    specimen_type_display == "MRSA Screen" ~ 
      "697989009", # Anterior nares swab 
    specimen_type_display == "Urine" ~ 
      "122575003", # Urine specimen
    TRUE ~ NA_character_
  )) %>% 
  left_join(transmute(reference_specimen_type,
                      specimen_type_code = conceptId,
                      specimen_type_name = pt_term))

```

The three data frames can now be created.

```{r create_three_micro_dataframes}
microbiology_data$specimens <- microbiology_data$raw %>% 
  transmute(specimen_id,
            patient_id,
            status = "available",
            specimen_datetime,
            specimen_type_code,
            specimen_type_name,
            specimen_type_display) %>% 
  distinct() # Removing duplicates created by multiple isolates and susceptibility testing

microbiology_data$isolates <- microbiology_data$raw %>% 
  transmute(organism_id,
            specimen_id,
            patient_id,
            organism_code,
            organism_name,
            organism_display_name) %>% 
  filter(!is.na(organism_code)) %>%  # Remove no growth
  distinct() # Removing duplicates created by susceptibility testing

microbiology_data$susceptibilities <- microbiology_data$raw %>% 
  filter(!is.na(organism_code)) %>%  # Remove no growth
  transmute(organism_id,
            specimen_id,
            patient_id,
            organism_code,
            organism_name,
            organism_display_name,
            drug_id,
            drug_name,
            drug_display_name,
            sir_code,
            concept_code = NA_character_) %>% 
  distinct()

```


An optional step is to classify isolates by multidrug resistance phenotype [@Magiorakos2012]. The AMR package function \code{\link[AMR]{mdro}()} provides classification based on:

* international guidelines on microorganism intrinsic resistance
* records of susceptibility testing (RSI). 


In this vignette, we adopt intrinsic resistances as defined by EUCAST guidelines [@EUCASTExpertRules32]. The tidyr package is used to pivot susceptibility test results into wide format (antimicrobials arranged as columns).

```{r classify_resistance_isolates}
mdr_classification <- microbiology_data$susceptibilities  %>% 
  select(organism_id,
         organism_code,
         drug_id,
         sir_code) %>% 
  tidyr::pivot_wider(id_cols = c("organism_id", "organism_code"),
                     names_from = drug_id,
                     values_from = sir_code)

mdr_classification$multidrug_resistance <- 
  as.character(AMR::mdro(
    x = mdr_classification,
    guideline = "EUCAST3.2",
    col_mo = "organism_code"))

# Merging back into the isolates dataset
microbiology_data$isolates <- left_join(
  microbiology_data$isolates,
  select(mdr_classification, organism_id, multidrug_resistance)
)
```

The data may now be validated, and loaded.

```{r}
validate_microbiology(specimens = microbiology_data$specimens,
                      isolates = microbiology_data$isolates,
                      susceptibilities = microbiology_data$susceptibilities)
```

```{r}
load_inpatient_microbiology(
  conn = ramses_db,
  specimens = microbiology_data$specimens,
  isolates = microbiology_data$isolates,
  susceptibilities = microbiology_data$susceptibilities,
  overwrite = TRUE
)
```

## Clinical investigations

Finally, we import all other observation and test results, from basic vitals such
as respiration rate and blood pressure, to urine antigen test results. Analysis of 
such data requires the use of standard test vocabularies.

The international standard code system for laboratory data is LOINC 
(\link[Ramses]{reference_loinc}). A sample of important codes may be found in 
dataset `Ramses::reference_loinc`. Other observable entity concepts or evaluation procedure concepts may be imported from SNOMED-CT.


```{r show_inpatient_investigations}
data(inpatient_investigations)
str(inpatient_investigations)
```

One particular set of validation constraints is that units conform to the [Unified Code for Units of Measure (UCUM)](https://ucum.org). Units are implemented using the `units` package in R. Yet, many important units in medicine are not listed in the UCUM. For instance: 

- respiration rate (breaths/min)
- heart rate (beats/min)
- alkaline phosphatase in serum (international unit per liter notated U/L)

By default, these three custom units are passed to the `custom_units` argument of `validate_investigations()`. Any other custom unit in the dataset must be handled with this argument. 

```{r validate_invest}
validate_investigations(inpatient_investigations, 
                        custom_units = c("breaths", "beats", "U"))
```

Data may now be loaded into the database.


```{r load_invest}
load_inpatient_investigations(
  conn = ramses_db,
  investigations_data = inpatient_investigations, 
  overwrite = TRUE
)
```


# Visualisation

You are now able to query the database and visualise admissions and episodes of antimicrobial therapy.

```{r}
therapy_timeline(conn = ramses_db, 
                 patient_identifier =  "99999999999",
                 date1 = as.Date("2017-02-01"),
                 date2 = as.Date("2017-03-01"))
```


# Once you are done 

Always remember close database connections when you are finished.

```{r}
DBI::dbDisconnect(ramses_db)
```



# References