---
title: "Ramses: R Package for Antimicrobial Stewardship & Surveillance"
date: '`r format(Sys.Date(), "%d %B %Y")`'
output: 
  rmarkdown::html_vignette:
    toc: true
    always_allow_html: yes
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{Ramses: R Package for Antimicrobial Stewardship & Surveillance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=6, 
  fig.height=4,
  fig.align = "center",
  df_print = "tibble"
)
```

<style type="text/css">
.blob {
  background-color: #193747;
  color: white;
  border-radius: 60px;
  min-inline-size: 60px;
  line-height: 60px;
  text-align: center;
  max-width:60px;
  margin: 1em;
  font-size: large;
  padding: 0pt;
}
}
.vcenter {
    display: inline-block;
    vertical-align: middle;
    float: none;
}
</style>

# Introduction

## Antimicrobial stewardship

Antimicrobial resistance jeopardizes many achievements of modern medicine, such as the ability to effectively treat infectious diseases or complete surgical procedures safely. Hospitals, alongside other inpatient facilities, make frequent use of categories of antibiotics which are most imminently eroding in effectiveness, the loss of which would have the most severe impact on health care systems [@Sharland2018]. They are thus at the forefront of the global effort to optimise the use of such antibiotics and sustain their effectiveness.

`Ramses` is an R package designed to support the stewardship of antimicrobials in hospitals by facilitating the analysis of routinely-collected electronic health records for:



<div class="container-fluid">
<div class="row">
  <div class="col-sm-6"><div class="col-sm-1 blob vcenter">ADT</div><div class="col-sm-8 vcenter">admission, discharge, transfers</div></div>
  <div class="col-sm-6"><div class="col-sm-1 blob vcenter">Rx</div><div class="col-sm-8 vcenter">drug prescribing and administrations</div></div>
  <div class="col-sm-6"><div class="col-sm-1 blob vcenter">Cx</div><div class="col-sm-8 vcenter">microbial culture and susceptibility</div></div>
  <div class="col-sm-6"><div class="col-sm-1 blob vcenter">Dx</div><div class="col-sm-8 vcenter">diagnostics, diagnoses and infection syndromes</div></div>
  <div class="col-sm-6"><div class="col-sm-1 blob vcenter">Ix</div><div class="col-sm-8 vcenter">clinical investigations and observations</div></div>
</div>
</div>

<!-- # Features -->


# Getting started 


## Setup 

`Ramses` is currently in development. The latest stable version can be installed using the `devtools` package.

```{r install ramses, eval=FALSE}
devtools::install_github("ramses-antibiotics/ramses-package")
```

`Ramses` comes with synthetic data for testing and training purposes.
You can use it to create a demonstration SQLite database fully loaded with 
synthetic electronic health records. 

```{r setup_mock_db}
library(Ramses)
library(ggplot2)
library(dplyr)
ramses_db <- create_mock_database("ramses-db.sqlite")
```

## Visualisation

You are now able to query the database and visualise admissions and episodes of antimicrobial therapy timeline below.

When hovered for a few seconds, timeline symbols display a tooltip containing detailed information.

```{r show_timeline, out.width = '100%'}
therapy_timeline(Patient(conn = ramses_db, 
                         id = "99999999999"),
                 date1 = as.Date("2017-02-01"),
                 date2 = as.Date("2017-03-01"))
```

## Data model

```{r list_tables}
DBI::dbListTables(ramses_db)
```


## Antibiotic consumption

### Overview

Three main metrics are used to describe antimicrobial consumption in hospitals: Defined Daily Doses, Days on Therapy, and Length of Therapy. Definitions given by @StanicBenic2018 are reproduced below:

- **Defined Daily Doses (DDDs)** designate "the assumed average maintenance dose per day for a drug used for its main indication in adults" as defined by the WHO Collaborating Centre for Drug Statistics Methodology (@WHO-ATC2020).
- **Days of Therapy (DOTs)** designate	"the number of days that a patient receives antibiotics regardless of the dose. When a patient receives more than one antibiotic, more than one DOT may be counted".
- **Length of Therapy (LOT)** designate "the number of days that a patient receives systemic antimicrobial agents, irrespective of the number of different antibiotics. Therefore, LOT will be lower than or equal to days of therapy (DOT) because the DOT is calculated for each antibiotic".

Formulae are available from @Ibrahim2014 along with a discussion of rates which can be derived from these metrics (eg per 1,000 admissions, per 1,000 bed-days). Statistical adjustment of these metrics is discussed by @VanSanten2018 and @Yu2018.

The approach to create tables of antimicrobial consumption follows three steps:

1. define the study population (encounters for which the above metrics should be derived)
2. choose a bridge table to link encounters with prescriptions
3. compute the metrics by aggregation.


### DOTs/DDDs based on the episode of administration 

Consumption can be attributed to specialty based on the specialty of the episode during which the antibiotic is administered. In this case, the appropriate bridge table is `bridge_episode_prescription_overlap`.

```{r AC1, paged.print=FALSE}
study_pop <- tbl(ramses_db, "inpatient_episodes") %>% 
  filter(main_specialty_code %in% c("100", "101", "300") &
           discharge_date >= "2016-01-01") %>% 
  mutate(calendar_year = year(discharge_date),
         calendar_month = month(discharge_date))  

consumption_episodes <- study_pop %>% 
  left_join(tbl(ramses_db, "bridge_episode_prescription_overlap")) %>% 
  group_by(calendar_year, calendar_month, main_specialty_code) %>% 
  summarise(DOT_prescribed = sum(DOT, na.rm = TRUE),
            DDD_prescribed = sum(DDD_prescribed, na.rm = TRUE),
            bed_days = sum(ramses_bed_days, na.rm = TRUE)) %>% 
  collect()  %>% 
  mutate(month_starting = as.Date(paste0(calendar_year, "/", calendar_month, "/01")))

head(consumption_episodes)
```

```{r plot_AC1,  width = 10}
ggplot(consumption_episodes, 
       aes(x = month_starting,
           y = DOT_prescribed/bed_days*1000, 
           group = main_specialty_code,
           color = main_specialty_code)) +
  geom_line() +
  scale_x_date(date_labels = "%b %Y") +
  xlab("Month") +
  ylab("Days of Therapy (DOTs)\nper 1,000 bed-days")
```


### DOTs/DDDs based on the episode of initiation 

Alternatively, consumption can be attributed to the episode when prescriptions are issued (`authoring_date` field in `drug_prescriptions`). In this case, the appropriate bridge table is `bridge_episode_prescription_initiation`. This amounts to attributing antibiotic consumption to the initial prescriber.
 

```{r AC2}
consumption_initiation <- study_pop %>% 
  left_join(tbl(ramses_db, "bridge_episode_prescription_initiation")) %>% 
  group_by(calendar_year, calendar_month, main_specialty_code) %>% 
  summarise(DOT_prescribed = sum(DOT, na.rm = TRUE),
            DDD_prescribed = sum(DDD_prescribed, na.rm = TRUE),
            total_admissions = n_distinct(paste0(patient_id, spell_id))) %>% 
  collect()  %>% 
  mutate(month_starting = as.Date(paste0(calendar_year, "/", calendar_month, "/01")))
```

```{r plot_AC2}
ggplot(consumption_initiation, 
       aes(x = month_starting,
           y = DOT_prescribed/total_admissions*1000, 
           group = main_specialty_code,
           color = main_specialty_code)) +
  geom_line() +
  scale_x_date(date_labels = "%b %Y") +
  xlab("Month") +
  ylab("Days of Therapy (DOTs)\nper 1,000 admissions")
```

### Length of therapy

Length of Therapy is the time elapsed during a prescribing episodes (sequence of antimicrobial prescriptions separated by at the most 36 hours by default). To measure it, the bridge table `bridge_spell_therapy_overlap` is available. It calculate the total LOT during between admission and discharge (excluding to-take-home medications).

```{r message=FALSE, results='asis'}
length_therapy_by_spell <- study_pop %>% 
  distinct(patient_id, spell_id, admission_method) %>% 
  left_join(tbl(ramses_db, "bridge_spell_therapy_overlap")) %>% 
  group_by(patient_id, spell_id, admission_method) %>% 
  summarise(LOT = sum(LOT, na.rm = T)) %>% 
  collect()

length_therapy_by_spell %>% 
  group_by(admission_method) %>% 
  summarise(
    `Total admissions` = n(),
    `Total admissions with therapy` = sum(!is.na(LOT)),
    `Mean LOT` = mean(LOT, na.rm = T),
    `Median LOT` = median(LOT, na.rm = T),
    percentile25 = quantile(x = LOT, probs = .25, na.rm = T),
    percentile75 = quantile(x = LOT, probs = .75, na.rm = T)
    ) %>% 
  transmute(
    `Admission method` = case_when(
      admission_method == 1 ~ "Elective",
      admission_method == 2 ~ "Emergency"),
    `Total admissions`,
    `Mean LOT`,
    `Median LOT`,
    `Inter-quartile range` = paste0(
      formatC(percentile25, format = "f", digits = 1),
      "-",
      formatC(percentile75, format = "f", digits = 1)
    )
  ) %>% 
  knitr::kable(digits = 1)
```


### Consumption based on the type of antibiotic

To measure rates of prescribing by antibiotic class or AWaRe category, the query design is different because different dimensions are sought for the numerator and denominator.

Unlike before, this task takes three steps:

1. compute numerator table
2. compute denominator table
3. join tables

```{r}
consumption_aware_episodes_num <- tbl(ramses_db, "drug_prescriptions") %>% 
  select(patient_id, prescription_id, ATC_code, ATC_route) %>% 
  left_join(tbl(ramses_db, "reference_aware"), 
            by = c("ATC_code", "ATC_route")) %>% 
  select(patient_id, prescription_id, aware_category) %>% 
  left_join(tbl(ramses_db, "bridge_episode_prescription_overlap")) %>% 
  inner_join(study_pop) %>% 
  group_by(calendar_year, calendar_month, aware_category) %>% 
  summarise(DOT_prescribed = sum(DOT, na.rm = TRUE)) %>% 
  collect()
 
consumption_aware_episodes_denom <- study_pop %>% 
  group_by(calendar_year, calendar_month) %>% 
  summarise(bed_days =  sum(ramses_bed_days, na.rm = TRUE)) %>% 
  collect()

consumption_aware_episodes <- left_join(
  consumption_aware_episodes_denom,
  consumption_aware_episodes_num
) %>% 
  mutate(month_starting = as.Date(paste0(calendar_year, "/", calendar_month, "/01")),
         aware_category = factor(aware_category, 
                                    levels = c("Access", "Watch", "Reserve")))

```

```{r}
aware_colours <- c(
  "Access" = "#1cb1d1", 
  "Watch" = "#008ab1", 
  "Reserve" = "#ff9667"
  )

ggplot(consumption_aware_episodes, 
       aes(x = month_starting,
           y = DOT_prescribed/bed_days*1000, 
           group = aware_category,
           color = aware_category)) +
  geom_line() +
  scale_x_date(date_labels = "%b %Y") +
  scale_color_manual(name = "AWaRe category", 
                     values = aware_colours) +
  xlab("Month") +
  ylab("Days of Therapy (DOTs)\nper 1,000 bed-days")
```





# Once you are done 

Always remember close database connections when you are finished.

```{r dbDisconnect}
DBI::dbDisconnect(ramses_db)
```


```{r include=FALSE}
file.remove("ramses-db.sqlite")
```


```{r eval=FALSE, include=FALSE}
# Do by Aware Rating next
# rx_atc_codes <- tbl(ramses_db, "drug_prescriptions") %>% 
#   select(prescription_id, ATC_code, ATC_route)
#   left_join(rx_atc_codes, by = "prescription_id") %>% 
#   left_join(tbl(ramses_db, "reference_aware"), by = c("ATC_code", "ATC_route")) %>% 
```

# References