---
title: "Ramses: R Package for Antimicrobial Stewardship & Surveillance"
output: 
  rmarkdown::html_vignette:
    toc: true
bibliography: ../inst/REFERENCES.bib
always_allow_html: yes
vignette: >
  %\VignetteIndexEntry{Ramses: R Package for Antimicrobial Stewardship & Surveillance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

<style type="text/css">
.blob {
  background-color: #193747;
  color: white;
  border-radius: 60px;
  min-inline-size: 60px;
  line-height: 60px;
  text-align: center;
  max-width:60px;
  margin: 1em;
}
.vcenter {
    display: inline-block;
    vertical-align: middle;
    float: none;
}
</style>

## Introduction

Antibiotic resistance jeopardizes many achievements of modern medicine, such as the ability to effectively treat infectious diseases or complete surgical procedures safely. Hospitals, alongside other inpatient facilities, make frequent use of categories of antibiotics which are most imminently erroding in effectiveness, the loss of which would have the most severe impact on health care systems [@Sharland2018]. They are thus at the forefront of the global effort to optimise the use of such antibiotics and sustain their effectiveness.

`Ramses` is an R package designed to support the stewardship of antimicrobials in hospitals by facilitating the analysis of routinely-collected electronic health records for:



<div class="container-fluid">
<div class="row">
  <div class="col-sm-6"><div class="col-sm-1 blob vcenter">ADT</div><div class="col-sm-8 vcenter">admission, discharge, transfers</div></div>
  <div class="col-sm-6"><div class="col-sm-1 blob vcenter">Rx</div><div class="col-sm-8 vcenter">drug prescribing and administration</div></div>
  <div class="col-sm-6"><div class="col-sm-1 blob vcenter">Cx</div><div class="col-sm-8 vcenter">microbial culture and susceptibility</div></div>
  <div class="col-sm-6"><div class="col-sm-1 blob vcenter">Dx</div><div class="col-sm-8 vcenter">diagnostics, diagnoses and infection syndromes</div></div>
  <div class="col-sm-6"><div class="col-sm-1 blob vcenter">Ix</div><div class="col-sm-8 vcenter">clinical investigations and observation.</div></div>
</div>
</div>

## Features

## Getting started 

```{r setup}
# library(Ramses)
```

## Reference