---
title: "Objects and classes in Ramses"
date: '`r format(Sys.Date(), "%d %B %Y")`'
output: 
  rmarkdown::html_vignette:
    toc: true 
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{Objects and classes in Ramses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(DiagrammeR)
# library(Ramses)
```

# Introduction and rationale

This article introduces the way `Ramses` handles entities such as patients, prescriptions, or therapy episodes. Each of these entities is associated with specific data analysis operations, and is represented in their own type of records in the backend database. 

To ensure users can access the relevant records and perform the desired analytical operations, `Ramses` employs a *object model* that consists of S4 classes representing each of these entities, and giving access to relevant functions or methods.

More information on, objects, classes, and methods in `R` is available from the [Object-oriented programming](https://adv-r.hadley.nz/oo.html) section by @Hadley2019.

# Objects in Ramses

`Ramses` objects are S4 classes. They all conform to the `RamsesObject` virtual class specification:

- they contain an `id` slot corresponding to a primary key in the database backend
- they contain a `conn` slot corresponding to the database backend connection, currently either of `duckdb_connection` or `PqConnection` class (see [database objects](#database-objects))
- they can be passed to `compute()` in order to compute the corresponding database records (faster to retrieve)
- they can be passed to `collect()` in order to retrieve the corresponding database records (returns a `tbl_df` object)
- they can be passed to `Patient()` in order to retrieve the corresponding patient entity (returns a `Patient` object).

`MedicationRequest` objects, which represent a single drug-dose order, which may or may not participate in combination therapy (2 or more drugs). `MedicationRequest` objects are associated with one episode of antimicrobial therapy (represented by `TherapyEpisode` objects). It is possible to retrieve the `TherapyEpisode` entity from any `MedicationRequest` object `x` by passing it to the `TherapyEpisode()` function.

<!--
`Ramses` distinguishes between:

- `MedicationRequest` objects, which represent a single drug-dose order
- `Prescription` objects, which may represent one or more drug-dose orders, depending on whether the patient is on monotherapy (single drug) or combination therapy (2 or more drugs).

Unlike `MedicationRequest` objects, `Prescription` objects may not be created directly with an `id`. `Prescription` objects must be created from a `MedicationRequest` object.

Both can be used to get to the episode of antimicrobial therapy (class: `TherapyEpisode`).
-->


```{r diagram-objects, fig.cap="Class diagram of the main `Ramses` objects", echo=FALSE}
htmltools::HTML(DiagrammeRsvg::export_svg(grViz("ramses-objects/ramses-class-diagram.gv")))
```

# Database objects

`Ramses` uses other objects for backend database connections and tables. They follow the [`dbplyr`](https://dbplyr.tidyverse.org/articles/dbplyr.html) architecture:

- database connections have class `DBIConnection`. Currently, `duckdb_connection` (package duckdb) and `PqConnection` (package RPostgres) are supported
- database table are created using the `tbl()` function and are represented by `tbl_sql` objects which, although they inherit from the `tbl` class, are distinct from the common `tbl_df` tibble class.

```{r diagram-tbl, fig.cap="Class diagram of tbl objects in the `tidyverse`", echo=FALSE}
htmltools::HTML(DiagrammeRsvg::export_svg(grViz("ramses-objects/tbl-class-diagram.gv")))
```


# References

