---
title: "Ramses: R Package for Antimicrobial Stewardship & Surveillance"
output: 
  rmarkdown::html_vignette:
    toc: true
bibliography: ../inst/REFERENCES.bib
always_allow_html: yes
vignette: >
  %\VignetteIndexEntry{Ramses: R Package for Antimicrobial Stewardship & Surveillance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

<style type="text/css">
.blob {
  background-color: #193747;
  color: white;
  border-radius: 60px;
  min-inline-size: 60px;
  line-height: 60px;
  text-align: center;
  max-width:60px;
  margin: 1em;
}
.vcenter {
    display: inline-block;
    vertical-align: middle;
    float: none;
}
</style>

# Introduction

## Antimicrobial stewardship

Antimicrobial resistance jeopardizes many achievements of modern medicine, such as the ability to effectively treat infectious diseases or complete surgical procedures safely. Hospitals, alongside other inpatient facilities, make frequent use of categories of antibiotics which are most imminently eroding in effectiveness, the loss of which would have the most severe impact on health care systems [@Sharland2018]. They are thus at the forefront of the global effort to optimise the use of such antibiotics and sustain their effectiveness.

`Ramses` is an R package designed to support the stewardship of antimicrobials in hospitals by facilitating the analysis of routinely-collected electronic health records for:



<div class="container-fluid">
<div class="row">
  <div class="col-sm-6"><div class="col-sm-1 blob vcenter">ADT</div><div class="col-sm-8 vcenter">admission, discharge, transfers</div></div>
  <div class="col-sm-6"><div class="col-sm-1 blob vcenter">Rx</div><div class="col-sm-8 vcenter">drug prescribing and administrations</div></div>
  <div class="col-sm-6"><div class="col-sm-1 blob vcenter">Cx</div><div class="col-sm-8 vcenter">microbial culture and susceptibility</div></div>
  <div class="col-sm-6"><div class="col-sm-1 blob vcenter">Dx</div><div class="col-sm-8 vcenter">diagnostics, diagnoses and infection syndromes</div></div>
  <div class="col-sm-6"><div class="col-sm-1 blob vcenter">Ix</div><div class="col-sm-8 vcenter">clinical investigations and observations</div></div>
</div>
</div>

# Features


# Getting started 


## Setup 

\code{Ramses} is currently in development. The latest stable version can be installed
using the \code{devtools} package.

```{r install ramses, eval=FALSE}
devtools::install_github("ramses-antibiotics/ramses-package")
```

\code{Ramses} comes with synthetic data for testing and training purposes.
You can use it to create a demonstration SQlite database fully loaded with 
synthectic electronic health records. 

```{r setup_mock_db}
library(Ramses)
ramses_db <- create_mock_database("ramses-db.sqlite")
```

## Visualisation

You are now able to query the database and visualise admissions and episodes of antimicrobial therapy timeline below.

When hovered for a few seconds, timeline symbols display a tooltip containing detailed information.

```{r show_timeline}
therapy_timeline(conn = ramses_db, 
                 patient_identifier = "99999999999",
                 date1 = as.Date("2017-02-01"),
                 date2 = as.Date("2017-03-01"))
```



# Once you are done 

Always remember close database connections when you are finished.

```{r}
DBI::dbDisconnect(ramses_db)
```


```{r include=FALSE}
file.remove("ramses-db.sqlite")
```

## Reference